---
title: "Lab 1: Intro to Quarto and Tidyverse"
author: "TJ Ayoub"
date: 11/01/23
date-format: "DD/MM/YY"
format: 
    pdf:
      toc: false
---

# Lab Exercises

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", 
                 skip = 2, col_types = "dcddd")
```

## 1. Plot the ratio of male to female mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme 

```{r}
p1 = dm |>
      filter(Age==10 | Age==20 | Age==30 | Age==40) |>
      mutate(mf_ratio = Male/Female) |>
      ggplot(aes(x=Year, y=mf_ratio, color=Age)) + geom_line() 

p1 + ylab("M:F Mortality Rate Ratio") + theme_minimal()
```

## 2. Find the age that has the highest female mortality rate each year 

```{r}
two = dm |>
        group_by(Year) |>
        slice_max(Female) |>
        select(Year:Age)

# Only a snippet shown for the first 10 years
two[1:10,]
```

## 3. Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations. 

```{r}
three = dm |>
        group_by(Age) |>
        summarize(across(c("Male","Female","Total"), sd))

three[1:10,]
```


## 4. The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here. 

```{r}
dp <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", 
                 skip = 2, col_types = "dcddd")
```

```{r}
dj = left_join(dm, dp, by = c("Year","Age"))

four = dj |>
        mutate(weightedMortFem = Female.x * Female.y / (Female.y + Male.y), 
               weightedMortMale = Male.x * Male.y / (Female.y + Male.y)) |>
        group_by(Year) |>
        summarize(popWeightFem = mean(weightedMortFem, na.rm=TRUE), 
                  popWeightMale = mean(weightedMortMale, na.rm=TRUE))

four[1:10,]
```

```{r}
four |> 
  pivot_longer(popWeightFem:popWeightMale, names_to = "Sex", values_to = "PWAMR") |>
  ggplot(aes(x=Year, y=PWAMR, color=Sex)) + geom_line() + 
    ylab("Population Weighted Average Mortality Rate") + 
    ggtitle("Change in PWAMR from 1921 to 2019") + 
    scale_color_hue(labels=c('Female', 'Male'))
```

We see a clear divergence in the average mortality rate when taking into account population size, between males and females.

